---
title: "V-Analysis"
output: html_document
date: "2025-04-17"
---

Vaishali ANOVA and T-test script



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Normalized the data based on FlowCut-passed/Cells/Single Cells/Live Cells|count.

Then based on the observations, "Age at Enrollment", "Sex", "Diabetes", "Athero", etc. I manipulated the data to prepare it for a t test or Anova test to gather significance. Based on the resulted meeting the 0.05 p value threshold 
those significant variable were then gathered and placed into a Box plot showing the data points of those significant variables in reference to the main observation. Lastly we loged the change in means and conducted Tukey test when necessary to make more significant claims on the observations and how the significant variables impacted them.

1. Age at enrollment new data frame (merged_data_norm_age)(above 60 and below 60)(Working On) Chart Outputted
2. Sex(women/men) new data frame(merged_data_norm_sex) (Working On) Chart Outputted
3. Athro new data frame(merged_data_norm_athro) (cabg or no cabg) If the column has (CABG or Coronary artery Bypass Graft) Label CAD else No-CAD
4. Diabetes new data frame(merged_data_norm_diabetes) (Working On) Chart Outputted
5. Smoking new data frame(merged_data_norm_smoking) ? (Working On) current smoker or not?just to 
6. Hyepertension new data frame(merged_data_norm_hypertension) (Working On) (Working On) Chart Outputted
7. BMI new data frame(merged_data_norm_bmi) (Buckets [Underweight BMI <18.5], [Average 18.5 - 24.9], [Overweight BMI 25 - 29.9], [Obese BMI >30]) (Working On) Chart Outputted
8. Hyperlipidemia new data frame(merged_data_norm_hyperlipidemia) (Working On)(Working On) Chart Outputted



Hyperlipidemia + hypertension
Seperate Age and BMI importance

CAD vs not CAD
Athro ( anything with CABG) or not

All Graphs
Check all Y axis




Normalizing the Data
```{r}
library(readr)
library(dplyr)

library(readr)
library(dplyr)

# Load the data
#merged_data_norm <- read_csv("/Users/jimkellypercine/Desktop/Merged_Flow_Data.csv")
merged_data_norm <- read_csv("/Users/ritikajain/Desktop/v_flow/data/Merged_Flow_Data.csv")


# Set the exact reference column name
ref_col <- "FlowCut-passed/Cells/Single Cells/Live Cells|count" 

# Normalize all child columns under Live Cells IN PLACE
merged_data_norm <- merged_data_norm %>%
  mutate(across(
    .cols = where(is.numeric) & contains("FlowCut-passed/Cells/Single Cells/Live Cells/") & 
            ends_with("|count") & 
            !all_of(ref_col),
    .fns = ~ .x / .data[[ref_col]]
  ))

# Check the result
head(merged_data_norm,50)
```



Observation (smoking)

```{r}
library(dplyr)
library(broom)
library(purrr)
library(stringr)
library(car)

# Create smoking_group column
merged_data_norm_smoking <- merged_data_norm %>%
  mutate(
    smoking_group = factor(
      if_else(`Current smoker` == "Yes", "Smoker", "Non-smoker"),
      levels = c("Non-smoker", "Smoker")
    )
  ) %>%
  arrange(smoking_group) %>%
  select(smoking_group, everything())

# Columns to test (numeric only)
smoking_cols_to_test <- merged_data_norm_smoking %>%
  select(where(is.numeric)) %>%
  colnames()

# Initialize p-value table
smoking_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# Run t-tests with Levene's test for equal variance
for (col in smoking_cols_to_test) {
  group1 <- merged_data_norm_smoking[merged_data_norm_smoking$smoking_group == "Smoker", col][[1]]
  group2 <- merged_data_norm_smoking[merged_data_norm_smoking$smoking_group == "Non-smoker", col][[1]]

  test_data <- data.frame(
    value = c(group1, group2),
    group = rep(c("Smoker", "Non-smoker"), c(length(group1), length(group2)))
  )

  levene_result <- tryCatch(leveneTest(value ~ group, data = test_data), error = function(e) NULL)
  equal_var <- TRUE
  if (!is.null(levene_result)) {
    equal_var <- levene_result$`Pr(>F)`[1] > 0.05
  }

  test_result <- tryCatch(t.test(group1, group2, var.equal = equal_var), error = function(e) NULL)

  if (!is.null(test_result)) {
    smoking_p_values <- rbind(smoking_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    smoking_p_values <- rbind(smoking_p_values, data.frame(column = col, p_value = NA))
  }
}

# Add significance flag
smoking_p_values <- smoking_p_values %>%
  mutate(significant = p_value < 0.05)

# Print formatted p-values
smoking_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

```

Smoking Charts

```{r}

library(ggplot2)

library(dplyr)

library(tidyr)

library(stringr)


# 1. Get significant columns with p-values and significance stars

sig_smoking_vars <- smoking_p_values %>%

  filter(significant == TRUE) %>%

  arrange(p_value) %>%

  mutate(

    significance = case_when(

      p_value < 0.001 ~ "***",

      p_value < 0.01 ~ "**",

      p_value < 0.05 ~ "*"

    ),

    short_name = LETTERS[seq_len(n())],

    short_label = paste0(short_name, " ", significance)

  )

 

# 2. Simplify variable names

simplified_names <- sig_smoking_vars$column %>%

  str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%

  str_replace_all("\\|count", "") %>%

  str_replace_all("\\|freq_of_parent", "%")

sig_smoking_vars$simplified_name <- simplified_names

 

# 3. Create named vector for labeling

smoking_label_map <- setNames(sig_smoking_vars$short_label, sig_smoking_vars$column)

 

# 4. Long-format data for plotting

long_data_smoking <- merged_data_norm_smoking %>%

  select(smoking_group, all_of(sig_smoking_vars$column)) %>%

  pivot_longer(cols = -smoking_group, names_to = "variable", values_to = "value") %>%

  mutate(

    value = value * 100,

    short_label = smoking_label_map[match(variable, sig_smoking_vars$column)]

  )

 

# 5. Plot

smoking_plot <- ggplot(long_data_smoking, aes(x = smoking_group, y = value)) +

  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +

  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +

  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,

               fill = "black", color = "black") +

  facet_wrap(~ short_label, scales = "free_y", ncol = 5, labeller = label_wrap_gen(width = 20)) +

  labs(

    title = "Distribution of Cell Populations by Smoking Status",

    subtitle = "Significant differences in cell frequency between Smokers and Non-smokers",

    x = "Smoking Group",

    y = "Percentage of Live Cells (%)",

    caption = "A: Central Memory CD4+ T Cells"

) +

  theme_minimal() +

 theme(
  plot.caption = element_text(
    hjust = 0.5,          # Centered
    size = 10, 
    face = "italic", 
    color = "black"
  ),
  axis.text.x = element_text(angle = 12, hjust = 0.5),
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5),
  strip.background = element_rect(fill = "lightblue", color = NA),
  strip.text = element_text(size = 14, face = "bold"),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(color = "gray90", fill = NA),
  axis.title = element_text(size = 12),
  axis.text = element_text(size = 10),
  plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
)
 

# Show plot

print(smoking_plot)

 

# 6. Optional: Print t-tests

for (i in 1:nrow(sig_smoking_vars)) {

  sig_var <- sig_smoking_vars$column[i]

  short_name <- sig_smoking_vars$short_name[i]

  simplified <- sig_smoking_vars$simplified_name[i]

 

  cat("\nStatistical test for", short_name, ":", simplified, ":\n")

  t_test_result <- t.test(as.formula(paste0("`", sig_var, "` ~ smoking_group")),

                          data = merged_data_norm_smoking)

  print(t_test_result)

}

 

# 7. Variable Key

cat("\nVariable Key:\n")

for (i in 1:nrow(sig_smoking_vars)) {

  cat(paste0(

    sig_smoking_vars$short_label[i], ": ",

    sig_smoking_vars$simplified_name[i],

    " (p = ", sprintf("%.4g", sig_smoking_vars$p_value[i]), ")",

    "\n"

  ))

}

 

```

Observation (Hyperlipidemia)

```{r}
library(dplyr)
library(broom)

# Create hyperlipidemia_group column

merged_data_norm_hyperlipidemia <- merged_data_norm %>%
  mutate(hyperlipidemia_group = if_else(Hyperlipidemia == "Yes", "Hyperlipidemic", "Non-hyperlipidemic")) %>%
  arrange(hyperlipidemia_group) %>%
  select(hyperlipidemia_group, everything())

# Columns to test (numeric only)

hyperlipidemia_cols_to_test <- merged_data_norm_hyperlipidemia %>%
  select(where(is.numeric)) %>%
  #select(-`Age at enrollment`, -BMI) %>%  # <-- Exclude Age and BMI
  colnames()

# Initialize p-value table

hyperlipidemia_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# Run t-tests
for (col in hyperlipidemia_cols_to_test) {
  group1 <- merged_data_norm_hyperlipidemia[merged_data_norm_hyperlipidemia$hyperlipidemia_group == "Hyperlipidemic", col]
  group2 <- merged_data_norm_hyperlipidemia[merged_data_norm_hyperlipidemia$hyperlipidemia_group == "Non-hyperlipidemic", col]
  
  test_result <- tryCatch(t.test(group1, group2, var.equal = TRUE), error = function(e) NULL)
  
  if (!is.null(test_result)) {
    hyperlipidemia_p_values <- rbind(hyperlipidemia_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    hyperlipidemia_p_values <- rbind(hyperlipidemia_p_values, data.frame(column = col, p_value = NA))
  }
}

# Add significance flag
hyperlipidemia_p_values <- hyperlipidemia_p_values %>%
  mutate(significant = p_value < 0.05)

# Print formatted p-values
hyperlipidemia_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# Manual test verification

t.test(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ hyperlipidemia_group,
       data = merged_data_norm_hyperlipidemia,
       var.equal = TRUE)

# Append p-values as new row

hyperlipidemia_p_row <- merged_data_norm_hyperlipidemia[nrow(merged_data_norm_hyperlipidemia), ]

hyperlipidemia_p_row[1, ] <- NA
hyperlipidemia_p_row$hyperlipidemia_group <- "p-value"

for (col in hyperlipidemia_cols_to_test) {
  p_val <- hyperlipidemia_p_values$p_value[hyperlipidemia_p_values$column == col]
  if (length(p_val) > 0) {
    hyperlipidemia_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

#Add p-values as new row

merged_data_with_hyperlipidemia_p_values <- rbind(merged_data_norm_hyperlipidemia, hyperlipidemia_p_row)

cat("\nLast rows of the dataset with hyperlipidemia p-values row added:\n")
tail(merged_data_with_hyperlipidemia_p_values, 5)

```


Hyperlipidemia Charts

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)

# 1. Get the significant columns with p-values and significance stars
sig_hyperlipidemia_vars <- hyperlipidemia_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    short_name = LETTERS[seq_len(n())],
    short_label = paste0(short_name, " ", significance)
  )

# 2. Split cell and non-cell variables
sig_cell_vars <- sig_hyperlipidemia_vars %>%
  filter(str_detect(column, "FlowCut"))

sig_noncell_vars <- sig_hyperlipidemia_vars %>%
  filter(!str_detect(column, "FlowCut"))

# 3. Simplified display names for cell variables
sig_cell_vars <- sig_cell_vars %>%
  mutate(
    simplified_name = column %>%
      str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%
      str_replace_all("\\|count", "") %>%
      str_replace_all("\\|freq_of_parent", "%")
  )

# 4. For non-cell variables, just use column name as simplified_name
sig_noncell_vars <- sig_noncell_vars %>%
  mutate(simplified_name = column)

# 5. Create label maps
label_map_cell <- setNames(sig_cell_vars$short_label, sig_cell_vars$column)
label_map_noncell <- setNames(sig_noncell_vars$short_label, sig_noncell_vars$column)

# 6. Long-format data for cell variables (scale to %)
long_data_cell <- merged_data_norm_hyperlipidemia %>%
  select(hyperlipidemia_group, all_of(sig_cell_vars$column)) %>%
  pivot_longer(cols = -hyperlipidemia_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,
    short_label = label_map_cell[variable]
  )

# 7. Long-format data for non-cell variables (no scaling)
long_data_noncell <- merged_data_norm_hyperlipidemia %>%
  select(hyperlipidemia_group, all_of(sig_noncell_vars$column)) %>%
  pivot_longer(cols = -hyperlipidemia_group, names_to = "variable", values_to = "value") %>%
  mutate(
    short_label = label_map_noncell[variable]
  )

# 8. Plot: Cell variables
hyperlipidemia_cell_plot <- ggplot(long_data_cell, aes(x = hyperlipidemia_group, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "black", color = "black") +
  facet_wrap(~ short_label, scales = "free_y", ncol = 5) +
  labs(
    title = "Distribution of Cell Populations by Hyperlipidemia Status",
    subtitle = "Significant differences in cell frequency between Hyperlipidemic and Non-hyperlipidemic",
    x = "Hyperlipidemia Group",
    y = "Percentage of Live Cells (%)",
    caption = "B: Naive CD8+ T Cells\nC: CD8+ T Cells expressing CCR7\nE: Hematopoietic Stem/Progenitor Cells (non-T Cells)"
    
  ) +
  theme_minimal() +
  
  theme(
     plot.caption = element_text(
    hjust = 0.5,          # Centered
    size = 10, 
    face = "italic", 
    color = "black"
  ),
    axis.text.x = element_text(angle = 12, hjust = 0.5),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(fill = "lightblue", color = NA),
    strip.text = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )

# 9. Plot: Non-cell variables
hyperlipidemia_noncell_plot <- ggplot(long_data_noncell, aes(x = hyperlipidemia_group, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "black", color = "black") +
  facet_wrap(~ short_label, scales = "free_y", ncol = 2) +
  labs(
    title = "Distribution of Non-Cell Features by Hyperlipidemia Status",
    subtitle = "Significant differences in Age and BMI between Hyperlipidemic and Non-hyperlipidemic",
    x = "Hyperlipidemia Group",
    y = "Clinical Measurement",
    caption = paste(
      paste0(sig_noncell_vars$short_label, ": ",
             sig_noncell_vars$simplified_name, " (",
             sig_noncell_vars$column, ")"),
      collapse = "\n"
    )
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 12, hjust = 0.5),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(fill = "lightblue", color = NA),
    strip.text = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic")
  )

# 10. Show the plots
print(hyperlipidemia_cell_plot)
print(hyperlipidemia_noncell_plot)

# 11. Optional: Print t-tests
for (i in 1:nrow(sig_hyperlipidemia_vars)) {
  sig_var <- sig_hyperlipidemia_vars$column[i]
  short_name <- sig_hyperlipidemia_vars$short_name[i]
  simplified <- sig_hyperlipidemia_vars$simplified_name[i]

  cat("\nStatistical test for", short_name, ":", simplified, ":\n")
  t_test_result <- t.test(as.formula(paste0("`", sig_var, "` ~ hyperlipidemia_group")),
                          data = merged_data_norm_hyperlipidemia)
  print(t_test_result)
}

# 12. Variable key with p-values
# 10. Variable Key with p-values
cat("\n### 🔑 Variable Key: Hypertension (Significant Cell Variables)\n\n")

for (i in seq_len(nrow(sig_cell_vars))) {
  line <- paste0(
    sig_cell_vars$short_name[i], " ", sig_cell_vars$significance[i], 
    ": ", sig_cell_vars$simplified_name[i],
    " (p = ", sprintf("%.6f", sig_cell_vars$p_value[i]), ")"
  )
  cat(line, "\n")
}

cat("\n### 🔑 Variable Key: Hypertension (Significant Cell Variables)\n\n")

for (i in seq_len(nrow(sig_noncell_vars))) {
  line <- paste0(
    sig_noncell_vars$short_name[i], " ", sig_noncell_vars$significance[i], 
    ": ", sig_noncell_vars$simplified_name[i],
    " (p = ", sprintf("%.6f", sig_noncell_vars$p_value[i]), ")"
  )
  cat(line, "\n")
}



```


Observation (hypertension)

```{r}

library(dplyr)
library(broom)
# Create hypertension_group column

merged_data_norm_hypertension <- merged_data_norm %>%
  mutate(hypertension_group = if_else(Hypertension == "Yes", "Hypertensive", "Non-hypertensive")) %>%
  arrange(hypertension_group) %>%
  select(hypertension_group, everything())

##########################
# Columns to test (numeric only)

hypertension_cols_to_test <- merged_data_norm_hypertension %>%
  select(where(is.numeric)) %>%
  #select(-`Age at enrollment`) %>%  # <-- Exclude Age and BMI
  colnames()

#########################
# Initialize p-value table

hypertension_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# Run t-tests

for (col in hypertension_cols_to_test) {
  group1 <- merged_data_norm_hypertension[merged_data_norm_hypertension$hypertension_group == "Hypertensive", col]
  group2 <- merged_data_norm_hypertension[merged_data_norm_hypertension$hypertension_group == "Non-hypertensive", col]
  
  test_result <- tryCatch(t.test(group1, group2, var.equal = TRUE), error = function(e) NULL)
  
  if (!is.null(test_result)) {
    hypertension_p_values <- rbind(hypertension_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    hypertension_p_values <- rbind(hypertension_p_values, data.frame(column = col, p_value = NA))
  }
}


#Add significance flag

hypertension_p_values <- hypertension_p_values %>%
  mutate(significant = p_value < 0.05)

# Print formatted p-values

hypertension_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# Manual test verification

t.test(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ hypertension_group,
       data = merged_data_norm_hypertension,
       var.equal = TRUE)

# Append p-values as new row

hypertension_p_row <- merged_data_norm_hypertension[nrow(merged_data_norm_hypertension), ]
hypertension_p_row[1, ] <- NA
hypertension_p_row$hypertension_group <- "p-value"

for (col in hypertension_cols_to_test) {
  p_val <- hypertension_p_values$p_value[hypertension_p_values$column == col]
  if (length(p_val) > 0) {
    hypertension_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

merged_data_with_hypertension_p_values <- rbind(merged_data_norm_hypertension, hypertension_p_row)

cat("\nLast rows of the dataset with hypertension p-values row added:\n")

tail(merged_data_with_hypertension_p_values, 5)
```




Hypertension Charts
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)

# 1. Get significant hypertension variables
sig_hypertension_vars <- hypertension_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    short_name = LETTERS[seq_len(n())],
    short_label = paste0(short_name, " ", significance)
  )

# 2. Filter and simplify cell variables only
sig_cell_vars <- sig_hypertension_vars %>%
  filter(str_detect(column, "FlowCut")) %>%
  mutate(
    simplified_name = column %>%
      str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%
      str_replace_all("\\|count", "") %>%
      str_replace_all("\\|freq_of_parent", "%")
  )

# 3. Create label map
label_map_cell <- setNames(sig_cell_vars$short_label, sig_cell_vars$column)

# 4. Long-format cell data
long_data_cell <- merged_data_norm_hypertension %>%
  select(hypertension_group, all_of(sig_cell_vars$column)) %>%
  pivot_longer(cols = -hypertension_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,
    short_variable = label_map_cell[variable]
  )

# 5. Analyze value ranges for grouping
range_analysis_cell <- long_data_cell %>%
  group_by(short_variable) %>%
  summarise(
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE),
    range_span = max_val - min_val,
    .groups = 'drop'
  ) %>%
  mutate(
    range_category = case_when(
      max_val <= 15 ~ "1",
      max_val <= 35 ~ "3",
      TRUE ~ "2"
    )
  )

# 6. Add category back to long data
long_data_cell_categorized <- long_data_cell %>%
  left_join(range_analysis_cell %>% select(short_variable, range_category), 
            by = "short_variable")

# 7. Plotting function
create_range_plot_hypertension <- function(data, category, fill_color = "lightblue", ncol_val = 4) {
  subset <- data %>% filter(range_category == category)
  
  ggplot(subset, aes(x = hypertension_group, y = value)) +
    geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fill = fill_color) +
    geom_jitter(width = 0.2, alpha = 0.6, size = 1.5, color = "gray40") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
    facet_wrap(~ short_variable, scales = "free_y", ncol = ncol_val) +
    scale_y_continuous(labels = function(x) sprintf("%.1f", x)) +
    labs(
      title = paste("Distribution of Cell Populations by Hypertension Group:", category),
      subtitle = "Significant differences in frequency between Hypertensive and Non-hypertensive",
      x = "Hypertension Group",
      y = "Percentage of Live Cells (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 9),
      axis.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(size = 12, face = "bold"),
      strip.background = element_rect(fill = fill_color, color = "black", size = 0.5),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", size = 0.3),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      plot.margin = margin(15, 15, 15, 15),
      panel.spacing = unit(0.8, "lines")
    )
}

# 8. Generate and print the separate plots
categories_cell <- unique(long_data_cell_categorized$range_category)
for (category in categories_cell) {
  cat("PLOTTING:", category, "\n")
  plot <- create_range_plot_hypertension(long_data_cell_categorized, category)
  print(plot)
}
# 8. Display the plots


# 9. Optional: Run post-hoc t-tests
for (i in 1:nrow(sig_hypertension_vars)) {
  sig_var <- sig_hypertension_vars$column[i]
  short_name <- sig_hypertension_vars$short_name[i]
  simplified <- sig_hypertension_vars$simplified_name[i]
  
  cat("\nStatistical test for", short_name, ":", simplified, ":\n")
  t_test_result <- t.test(as.formula(paste0("`", sig_var, "` ~ hypertension_group")), data = merged_data_norm_hypertension)
  print(t_test_result)
}


# 10. Variable Key with p-values
cat("\n### 🔑 Variable Key: Hypertension (Significant Cell Variables)\n\n")

for (i in seq_len(nrow(sig_cell_vars))) {
  line <- paste0(
    sig_cell_vars$short_name[i], " ", sig_cell_vars$significance[i], 
    ": ", sig_cell_vars$simplified_name[i],
    " (p = ", sprintf("%.6f", sig_cell_vars$p_value[i]), ")"
  )
  cat(line, "\n")
}

```


Observation (diabetes)

```{r}
library(dplyr)
library(broom)

# Create diabetes_group column

merged_data_norm_diabetes <- merged_data_norm %>%
  mutate(diabetes_group = if_else(Diabetes == "Yes", "Diabetic", "Non-diabetic")) %>%
  arrange(diabetes_group) %>%
  select(diabetes_group, everything())

# Columns to test (numeric only)

diabetes_cols_to_test <- merged_data_norm_diabetes %>%
  select(where(is.numeric)) %>%
  colnames()

# Initialize p-value table

diabetes_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# Run t-tests

for (col in diabetes_cols_to_test) {
  group1 <- merged_data_norm_diabetes[merged_data_norm_diabetes$diabetes_group == "Diabetic", col]
  group2 <- merged_data_norm_diabetes[merged_data_norm_diabetes$diabetes_group == "Non-diabetic", col]
  
  test_result <- tryCatch(t.test(group1, group2, var.equal = TRUE), error = function(e) NULL)
  
  if (!is.null(test_result)) {
    diabetes_p_values <- rbind(diabetes_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    diabetes_p_values <- rbind(diabetes_p_values, data.frame(column = col, p_value = NA))
  }
}

# Add significance flag

diabetes_p_values <- diabetes_p_values %>%
  mutate(significant = p_value < 0.05)

# Print formatted p-values

diabetes_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# Manual test verification

t.test(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ diabetes_group,
       data = merged_data_norm_diabetes,
       var.equal = TRUE)

# Append p-values as new row

diabetes_p_row <- merged_data_norm_diabetes[nrow(merged_data_norm_diabetes), ]
diabetes_p_row[1, ] <- NA
diabetes_p_row$diabetes_group <- "p-value"

for (col in diabetes_cols_to_test) {
  p_val <- diabetes_p_values$p_value[diabetes_p_values$column == col]
  if (length(p_val) > 0) {
    diabetes_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

merged_data_with_diabetes_p_values <- rbind(merged_data_norm_diabetes, diabetes_p_row)

cat("\nLast rows of the dataset with diabetes p-values row added:\n")
tail(merged_data_with_diabetes_p_values, 5)
```


Diabetes charts
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)

# 1. Get significant variables from diabetes p-value results
sig_diabetes_vars <- diabetes_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    short_name = LETTERS[seq_len(n())],
    short_label = paste0(short_name, " ", significance)
  )

# 2. Create simplified variable names for display
sig_diabetes_vars$simplified_name <- sig_diabetes_vars$column %>%
  str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%
  str_replace_all("\\|count", "") %>%
  str_replace_all("\\|freq_of_parent", "%")

# 3. Prepare long-format data
long_data_diabetes <- merged_data_norm_diabetes %>%
  select(diabetes_group, all_of(sig_diabetes_vars$column)) %>%
  pivot_longer(cols = -diabetes_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,
    short_variable = sig_diabetes_vars$short_label[match(variable, sig_diabetes_vars$column)]
  )

# 4. Analyze ranges to categorize variables
range_analysis_diabetes <- long_data_diabetes %>%
  group_by(short_variable) %>%
  summarise(
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE),
    range_span = max_val - min_val,
    .groups = 'drop'
  ) %>%
  mutate(
    range_category = case_when(
      max_val <= 15 ~ "Distribution of Significant Variables by Diabetes Group 1",
      max_val <= 35 ~ "Distribution of Significant Variables by Diabetes Group 3",
      TRUE ~ "Distribution of Significant Variables by Diabetes Group 2"
    )
  )

# 5. Add range category to the main data
long_data_diabetes_categorized <- long_data_diabetes %>%
  left_join(
    range_analysis_diabetes %>% select(short_variable, range_category), 
    by = "short_variable"
  )

# 6. Create plotting function
create_range_plot_diabetes <- function(data, category, fill_color, ncol_val = 4) {
  data_subset <- data %>% filter(range_category == category)

  ggplot(data_subset, aes(x = diabetes_group, y = value)) +
    geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fill = fill_color) +
    geom_jitter(width = 0.2, alpha = 0.6, size = 1.5, color = "gray40") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
    facet_wrap(~ short_variable, scales = "free_y", ncol = ncol_val) +
    scale_y_continuous(labels = function(x) sprintf("%.1f", x)) +
    labs(
      title = paste(category),
      subtitle = paste("Significant differences by diabetes group"),
      x = "Diabetes Group",
      y = "Percentage of Live Cells (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 9),
      axis.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(size = 12, face = "bold"),
      strip.background = element_rect(fill = fill_color, color = "black", size = 0.5),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", size = 0.3),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      plot.margin = margin(15, 15, 15, 15),
      panel.spacing = unit(0.8, "lines")
    )
}

# 7. Plot each frequency category
categories_diabetes <- unique(long_data_diabetes_categorized$range_category)

for (category in categories_diabetes) {
  cat("PLOTTING:", category, "\n")
  fill_color <- "lightblue"
  ncol_val <- ifelse(grepl("Low", category), 4, 3)

  plot <- create_range_plot_diabetes(long_data_diabetes_categorized, category, fill_color, ncol_val)
  print(plot)
}

# 8. Create variable legend
variable_legend_diabetes <- sig_diabetes_vars %>%
  left_join(
    range_analysis_diabetes %>% select(short_variable, range_category),
    by = c("short_label" = "short_variable")
  ) %>%
  mutate(
    p_formatted = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01 ~ sprintf("%.3f", p_value),
      TRUE ~ sprintf("%.4f", p_value)
    )
  ) %>%
  select(short_name, significance, simplified_name, p_formatted, range_category) %>%
  arrange(range_category, short_name)

for (category in categories_diabetes) {
  cat("\n", category, ":\n")
  category_vars <- variable_legend_diabetes %>% filter(range_category == category)
  for (i in 1:nrow(category_vars)) {
    cat(sprintf("%s (%s): %s (p = %s)\n", 
                category_vars$short_name[i], 
                category_vars$significance[i],
                category_vars$simplified_name[i], 
                category_vars$p_formatted[i]))
  }
}

# 9. Summary statistics
summary_by_category_diabetes <- long_data_diabetes_categorized %>%
  group_by(range_category, short_variable, diabetes_group) %>%
  summarise(
    n = n(),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = 'drop'
  ) %>%
  arrange(range_category, short_variable, diabetes_group)

for (category in categories_diabetes) {
  cat("\n", category, ":\n")
  print(summary_by_category_diabetes %>% filter(range_category == category))
}

# 10. Final variable key
cat("\nVariable Key:\n")
for (i in 1:nrow(sig_diabetes_vars)) {
  cat(paste0(
    sig_diabetes_vars$short_label[i], ": ",
    sig_diabetes_vars$simplified_name[i],
    " (p = ", sprintf("%.4g", sig_diabetes_vars$p_value[i]), ")\n"
  ))
}

cat("\nANALYSIS COMPLETE\n")
cat("Variables have been separated into three charts based on frequency ranges:\n")
cat("\u2022 Low Frequency (0-15%): Rare cell populations\n")
cat("\u2022 Medium Frequency (15-35%): Moderate cell populations\n") 
cat("\u2022 High Frequency (35%+): Abundant cell populations\n")

```

Observation (age)
```{r}
library(dplyr)
library(broom)

# Create age_group column with buckets
merged_data_norm_age <- merged_data_norm %>%
  mutate(age_group = case_when(
    `Age at enrollment` >= 24 & `Age at enrollment` <= 59 ~ "24–59",
    `Age at enrollment` >= 60 & `Age at enrollment` <= 69 ~ "60–69",
    `Age at enrollment` >= 70 & `Age at enrollment` <= 89 ~ "70–89",
    TRUE ~ NA_character_
  )) %>%
  arrange(age_group) %>%
  select(age_group, everything())

# Columns to test (exclude Age at enrollment and non-numeric IDs)
age_cols_to_test <- merged_data_norm_age %>%
  select(where(is.numeric), -`Age at enrollment`, -`Record ID`) %>%
  colnames()

# Initialize p-value table
age_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# Run ANOVA tests
for (col in age_cols_to_test) {
  formula <- as.formula(paste0("`", col, "` ~ age_group"))
  aov_result <- tryCatch(aov(formula, data = merged_data_norm_age), error = function(e) NULL)
  
  if (!is.null(aov_result)) {
    p_val <- summary(aov_result)[[1]][["Pr(>F)"]][1]
    age_p_values <- rbind(age_p_values, data.frame(column = col, p_value = p_val))
  } else {
    age_p_values <- rbind(age_p_values, data.frame(column = col, p_value = NA))
  }
}

# Add significance flag
age_p_values <- age_p_values %>%
  mutate(significant = p_value < 0.05)

# Print formatted p-values
age_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# Optional manual check
summary(aov(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ age_group,
            data = merged_data_norm_age))

# Append p-values as a new row (for completeness)
age_p_row <- merged_data_norm_age[nrow(merged_data_norm_age), ]
age_p_row[1, ] <- NA
age_p_row$age_group <- "p-value"

for (col in age_cols_to_test) {
  p_val <- age_p_values$p_value[age_p_values$column == col]
  if (length(p_val) > 0) {
    age_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

merged_data_with_age_p_values <- rbind(merged_data_norm_age, age_p_row)


# 8. Optional: Run post-hoc Tukey tests for significant age variables
significant_age_vars <- age_p_values %>% filter(significant) %>% pull(column)

# Print Tukey results for each significant variable
for (var in significant_age_vars) {
  cat("\nPairwise comparisons for:", var, "\n")
  formula <- as.formula(paste0("`", var, "` ~ age_group"))
  aov_model <- aov(formula, data = merged_data_norm_age)
  tukey <- TukeyHSD(aov_model)
  print(tukey)
}

# 8.1 Create a summary table for Tukey's HSD results (all comparisons)
tukey_age_summary <- data.frame(
  variable = character(),
  comparison = character(),
  estimate = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (var in significant_age_vars) {
  formula <- as.formula(paste0("`", var, "` ~ age_group"))
  aov_model <- aov(formula, data = merged_data_norm_age)
  tukey_result <- TukeyHSD(aov_model)
  
  # Loop through comparisons and extract all results
  for (comparison in rownames(tukey_result[[1]])) {
    estimate <- tukey_result[[1]][comparison, "diff"]
    p_value <- tukey_result[[1]][comparison, "p adj"]
    
    tukey_age_summary <- rbind(tukey_age_summary, data.frame(
      variable = var,
      comparison = comparison,
      estimate = estimate,
      p_value = sprintf("%.4f", p_value)  # Format p-value to 4 decimal places
    ))
  }
}

# Optionally, filter for significant comparisons only (p < 0.05)
tukey_age_summary_sig <- tukey_age_summary %>% filter(p_value < 0.05)

# Print the summary table
cat("\nTukey HSD summary table for age_group (all comparisons):\n")
print(tukey_age_summary)

cat("\nTukey HSD summary table for age_group (significant only):\n")
print(tukey_age_summary_sig)


cat("\nLast rows of the dataset with age p-values row added:\n")
tail(merged_data_with_age_p_values, 5)
```


Observation (age charts)
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)

# 1. Get the significant columns with p-values and significance stars
sig_age_vars <- age_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    # Create short variable names for better readability
    short_name = LETTERS[seq_len(n())],
    short_label = paste0(short_name, " ", significance)
  )

# 2. Create simplified variable names for display
simplified_names <- sig_age_vars$column %>%
  str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%
  str_replace_all("\\|count", "") %>%
  str_replace_all("\\|freq_of_parent", "%") 

sig_age_vars$simplified_name <- simplified_names

# 3. Prepare long-format data
long_data_age <- merged_data_norm_age %>%
  select(age_group, all_of(sig_age_vars$column)) %>%
  pivot_longer(cols = -age_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,  # Convert to percentage
    short_variable = sig_age_vars$short_label[match(variable, sig_age_vars$column)]
  )

# 4. Analyze ranges to categorize variables
range_analysis <- long_data_age %>%
  group_by(short_variable) %>%
  summarise(
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE),
    range_span = max_val - min_val,
    .groups = 'drop'
  ) %>%
  mutate(
    range_category = case_when(
      max_val <= 15 ~ "Distribution of Significant Variables by Age Group 1",
      max_val <= 35 ~ "Distribution of Significant Variables by Age Group 3", 
      TRUE ~ "Distribution of Significant Variables by Age Group 2"
    )
  ) %>%
  arrange(max_val)

# Display range analysis
cat("Variable Range Categories:\n")
#cat("="*50, "\n")
print(range_analysis)

# 5. Add range category to the main data
long_data_age_categorized <- long_data_age %>%
  left_join(
    range_analysis %>% select(short_variable, range_category), 
    by = "short_variable"
  )

# 6. Create function to generate plots for each category
create_range_plot <- function(data, category, fill_color, ncol_val = 4) {
  data_subset <- data %>% filter(range_category == category)
  
  # Count variables in this category
  n_vars <- length(unique(data_subset$short_variable))
  
  ggplot(data_subset, aes(x = age_group, y = value)) +
    geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fill = fill_color) +
    geom_jitter(width = 0.2, alpha = 0.6, size = 1.5, color = "gray40") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
    facet_wrap(~ short_variable, scales = "free_y", ncol = ncol_val) +
    scale_y_continuous(labels = function(x) sprintf("%.1f", x)) +
    labs(
      title = paste(category),
      subtitle = paste("Significant differences by age group"),
      x = "Age Group",
      y = "Percentage of Live Cells (%)",
      caption = ""
    ) +
    theme_bw(base_size = 12) +
    theme(
      
      plot.caption = element_text(
    hjust = 0.5,          # Centered
    size = 10, 
    face = "italic", 
    color = "black"
  ),
      
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 9),
      axis.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(size = 12, face = "bold"),
      strip.background = element_rect(fill = fill_color, color = "black", size = 0.5),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", size = 0.3),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      plot.margin = margin(15, 15, 15, 15),
      panel.spacing = unit(0.8, "lines")
    )
}

# 7. Generate separate plots for each frequency category
categories <- unique(long_data_age_categorized$range_category)

for (category in categories) {
  #cat("\n", "="*60, "\n")
  cat("PLOTTING:", category, "\n")
  #cat("="*60, "\n")
  
  # Choose colors and layout based on category
  if (grepl("Low", category)) {
    fill_color <- "lightblue"
    ncol_val <- 4
  } else if (grepl("Medium", category)) {
    fill_color <- "lightblue"
    ncol_val <- 3
  } else {
    fill_color <- "lightblue"  
    ncol_val <- 3
  }
  
  # Create and print plot
  plot <- create_range_plot(long_data_age_categorized, category, fill_color, ncol_val)
  print(plot)
}

# 8. Create variable legend for each category
#cat("\n\n", "="*80, "\n")
cat("VARIABLE LEGEND BY CATEGORY\n")
#cat("="*80, "\n")

# Get variable details with categories
variable_legend <- sig_age_vars %>%
  left_join(
    range_analysis %>% select(short_variable = short_variable, range_category), 
    by = c("short_label" = "short_variable")
  ) %>%
  mutate(
    p_formatted = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01 ~ sprintf("%.3f", p_value),
      TRUE ~ sprintf("%.4f", p_value)
    )
  ) %>%
  select(short_name, significance, simplified_name, p_formatted, range_category) %>%
  arrange(range_category, short_name)

# Print legend by category
for (category in categories) {
  cat("\n", category, ":\n")
  #cat("-"*40, "\n")
  
  category_vars <- variable_legend %>% filter(range_category == category)
  
  for (i in 1:nrow(category_vars)) {
    cat(sprintf("%s (%s): %s (p = %s)\n", 
                category_vars$short_name[i], 
                category_vars$significance[i],
                category_vars$simplified_name[i], 
                category_vars$p_formatted[i]))
  }
}

# 9. Summary statistics by category
#cat("\n\n", "="*80, "\n")
cat("SUMMARY STATISTICS BY CATEGORY\n")
#cat("="*80, "\n")

summary_by_category <- long_data_age_categorized %>%
  group_by(range_category, short_variable, age_group) %>%
  summarise(
    n = n(),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = 'drop'
  ) %>%
  arrange(range_category, short_variable, age_group)

for (category in categories) {
  cat("\n", category, ":\n")
  #cat("-"*50, "\n")
  
  category_summary <- summary_by_category %>% filter(range_category == category)
  print(category_summary)
}


cat("\nVariable Key:\n")
for (i in 1:nrow(sig_age_vars)) {
  cat(paste0(
    sig_age_vars$short_label[i], ": ",
    sig_age_vars$simplified_name[i],
    " (p = ", sprintf("%.4g", sig_age_vars$p_value[i]), ")",
    "\n"
  ))
}

#cat("\n\n", "="*80, "\n")
cat("ANALYSIS COMPLETE\n")
#cat("="*80, "\n")
cat("Variables have been separated into three charts based on frequency ranges:\n")
cat("• Low Frequency (0-15%): Rare cell populations\n")
cat("• Medium Frequency (15-35%): Moderate cell populations\n") 
cat("• High Frequency (35%+): Abundant cell populations\n")
cat("\nThis approach makes it easier to:\n")
cat("• Compare variables within similar ranges\n")
cat("• Identify patterns specific to each frequency category\n")
cat("• Present results to different audiences\n")
#cat("="*80, "\n")
```

Observation Sex(Women/Men)
```{r}
# Create sex_group column
merged_data_norm_sex <- merged_data_norm %>%
  mutate(sex_group = if_else(`Sex assigned at birth` == "Female", "Women", "Men")) %>%
  arrange(sex_group) %>%
  select(sex_group, everything())

# Columns to test (numeric only)
sex_cols_to_test <- merged_data_norm_sex %>%
  select(where(is.numeric)) %>%
  colnames()

# Initialize p-value table
sex_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# Run t-tests
for (col in sex_cols_to_test) {
  group1 <- merged_data_norm_sex[merged_data_norm_sex$sex_group == "Women", col]
  group2 <- merged_data_norm_sex[merged_data_norm_sex$sex_group == "Men", col]
  
  test_result <- tryCatch(t.test(group1, group2, var.equal = TRUE), error = function(e) NULL)
  
  if (!is.null(test_result)) {
    sex_p_values <- rbind(sex_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    sex_p_values <- rbind(sex_p_values, data.frame(column = col, p_value = NA))
  }
}

# Add significance flag
sex_p_values <- sex_p_values %>%
  mutate(significant = p_value < 0.05)

# Print formatted p-values
sex_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# Manual test verification
t.test(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ sex_group,
       data = merged_data_norm_sex,
       var.equal = TRUE)

# Append p-values as new row
sex_p_row <- merged_data_norm_sex[nrow(merged_data_norm_sex), ]
sex_p_row[1, ] <- NA
sex_p_row$sex_group <- "p-value"

for (col in sex_cols_to_test) {
  p_val <- sex_p_values$p_value[sex_p_values$column == col]
  if (length(p_val) > 0) {
    sex_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

merged_data_with_sex_p_values <- rbind(merged_data_norm_sex, sex_p_row)

cat("\nLast rows of the dataset with sex p-values row added:\n")
tail(merged_data_with_sex_p_values, 5)
```

Sex(M/F) Charts
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(cowplot)

# 1. Get the significant columns with p-values and significance stars
sig_sex_vars <- sex_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    short_name = LETTERS[seq_len(n())],
    short_label = paste0(short_name, " ", significance)
  )

# 2. Create simplified variable names for display
simplified_names <- sig_sex_vars$column %>%
  str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%
  str_replace_all("\\|count", "") %>%
  str_replace_all("\\|freq_of_parent", "%") 

sig_sex_vars$simplified_name <- simplified_names

# 3. Prepare long-format data
long_data_sex <- merged_data_norm_sex %>%
  select(sex_group, all_of(sig_sex_vars$column)) %>%
  pivot_longer(cols = -sex_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,  # Convert to percentage
    short_variable = sig_sex_vars$short_label[match(variable, sig_sex_vars$column)]
  )

# 4. Analyze ranges to categorize variables
range_analysis <- long_data_sex %>%
  group_by(short_variable) %>%
  summarise(
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE),
    range_span = max_val - min_val,
    .groups = 'drop'
  ) %>%
  mutate(
    range_category = case_when(
      max_val <= 15 ~ "Distribution of Significant Variables by Sex Group 2",
      max_val <= 35 ~ "Distribution of Significant Variables by Sex Group 1", 
      TRUE ~ "Distribution of Significant Variables by Sex Group 3"
    )
  ) %>%
  arrange(max_val)

# Display range analysis
cat("Variable Range Categories:\n")
print(range_analysis)

# 5. Add range category to the main data
long_data_sex_categorized <- long_data_sex %>%
  left_join(
    range_analysis %>% select(short_variable, range_category), 
    by = "short_variable"
  )

# 6. Enhanced function to generate plots with variable names
create_sex_range_plot <- function(data, category, fill_color, ncol_val = 4) {
  data_subset <- data %>% filter(range_category == category)
  
  # Count variables in this category
  n_vars <- length(unique(data_subset$short_variable))
  
  # Get the variables for this category to create the legend
  category_vars <- sig_sex_vars %>%
    filter(short_label %in% unique(data_subset$short_variable)) %>%
    arrange(short_name)
  
  # Create the main plot
  main_plot <- ggplot(data_subset, aes(x = sex_group, y = value)) +
    geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fill = fill_color) +
    geom_jitter(width = 0.2, alpha = 0.6, size = 1.5, color = "gray40") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
    facet_wrap(~ short_variable, scales = "free_y", ncol = ncol_val) +
    scale_y_continuous(labels = function(x) sprintf("%.1f", x)) +
    labs(
      title = paste(category),
      subtitle = paste("Significant differences by sex group"),
      x = "Sex Group",
      y = "Percentage of Live Cells (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
      axis.text.y = element_text(size = 9),
      axis.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(size = 12, face = "bold"),
      strip.background = element_rect(fill = fill_color, color = "black", size = 0.5),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", size = 0.3),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      plot.margin = margin(15, 15, 15, 15),
      panel.spacing = unit(0.8, "lines")
    )
  
  # Create legend text as separate plot
  legend_lines <- paste0(category_vars$short_name, ": ", category_vars$simplified_name)
  
  # Format into columns based on number of variables
  n_vars <- length(legend_lines)
  if (n_vars > 6) {
    # Split into 2 columns for better readability
    mid_point <- ceiling(n_vars / 2)
    col1 <- legend_lines[1:mid_point]
    col2 <- legend_lines[(mid_point + 1):n_vars]
    
    # Pad shorter column
    max_len <- max(length(col1), length(col2))
    col1 <- c(col1, rep("", max_len - length(col1)))
    col2 <- c(col2, rep("", max_len - length(col2)))
    
    legend_text <- paste(
      paste(col1, collapse = "\n"),
      "     ",
      paste(col2, collapse = "\n"),
      sep = ""
    )
  } else {
    legend_text <- paste(legend_lines, collapse = "\n")
  }
  
  # Create text plot for legend with better formatting
  legend_plot <- ggplot() + 
    annotate("text", x = 0.02, y = 0.5, label = legend_text, 
             hjust = 0, vjust = 0.5, size = 3.5, 
             family = "", lineheight = 1.2) +
    xlim(0, 1) + ylim(0, 1) +
    theme_void() +
    theme(plot.margin = margin(10, 10, 10, 10))
  
  # Combine plots with more space for legend
  final_plot <- plot_grid(
    main_plot,
    legend_plot,
    ncol = 1,
    rel_heights = c(0.75, 0.25),
    align = "v"
  )
  
  return(final_plot)
}

# 7. Generate separate plots for each frequency category
categories <- unique(long_data_sex_categorized$range_category)

for (category in categories) {
  cat("PLOTTING:", category, "\n")
  
  # Choose colors and layout based on category
  if (grepl("1", category)) {
    fill_color <- "lightblue"
    ncol_val <- 4
  } else if (grepl("3", category)) {
    fill_color <- "lightblue"
    ncol_val <- 3
  } else {
    fill_color <- "lightblue"  
    ncol_val <- 3
  }
  
  # Create and print plot with variable names
  plot <- create_sex_range_plot(long_data_sex_categorized, category, fill_color, ncol_val)
  print(plot)
}

# 8. Create variable legend for each category
cat("VARIABLE LEGEND BY CATEGORY\n")

# Get variable details with categories
variable_legend <- sig_sex_vars %>%
  left_join(
    range_analysis %>% select(short_variable = short_variable, range_category), 
    by = c("short_label" = "short_variable")
  ) %>%
  mutate(
    p_formatted = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01 ~ sprintf("%.3f", p_value),
      TRUE ~ sprintf("%.4f", p_value)
    )
  ) %>%
  select(short_name, significance, simplified_name, p_formatted, range_category) %>%
  arrange(range_category, short_name)

# Print legend by category
for (category in categories) {
  cat("\n", category, ":\n")
  
  category_vars <- variable_legend %>% filter(range_category == category)
  
  for (i in 1:nrow(category_vars)) {
    cat(sprintf("%s (%s): %s (p = %s)\n", 
                category_vars$short_name[i], 
                category_vars$significance[i],
                category_vars$simplified_name[i], 
                category_vars$p_formatted[i]))
  }
}

# 9. Summary statistics by category
cat("SUMMARY STATISTICS BY CATEGORY\n")

summary_by_category <- long_data_sex_categorized %>%
  group_by(range_category, short_variable, sex_group) %>%
  summarise(
    n = n(),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = 'drop'
  ) %>%
  arrange(range_category, short_variable, sex_group)

for (category in categories) {
  cat("\n", category, ":\n")
  
  category_summary <- summary_by_category %>% filter(range_category == category)
  print(category_summary)
}


# Print variable key with p-values
cat("\nVariable Key:\n")
for (i in 1:nrow(sig_sex_vars)) {
  cat(paste0(
    sig_sex_vars$short_label[i], ": ",
    sig_sex_vars$simplified_name[i],
    " (p = ", sprintf("%.4g", sig_sex_vars$p_value[i]), ")",
    "\n"
  ))
}

```


Observation (BMI)
```{r}
library(dplyr)

# 1. Create bmi_group column and remove "Underweight" group
merged_data_norm_bmi <- merged_data_norm %>%
  mutate(bmi_group = case_when(
    BMI >= 18.5 & BMI <= 24.9 ~ "Average",
    BMI > 25 & BMI <= 29.9 ~ "Overweight",
    BMI > 30 ~ "Obese",
    TRUE ~ NA_character_
  )) %>%
  filter(BMI >= 18.5) %>%  # remove NA and Underweight
  arrange(BMI) %>%
  select(bmi_group, BMI, everything())

# 2. Select numeric columns to test (excluding BMI itself)
bmi_cols_to_test <- merged_data_norm_bmi %>%
  select(where(is.numeric), -BMI) %>%
  colnames()

# 3. Run one-way ANOVA and collect p-values
bmi_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

for (col in bmi_cols_to_test) {
  test_result <- tryCatch({
    formula <- as.formula(paste0("`", col, "` ~ bmi_group"))
    aov_result <- aov(formula, data = merged_data_norm_bmi)
    summary(aov_result)[[1]][["Pr(>F)"]][1]
  }, error = function(e) NA)
  
  bmi_p_values <- rbind(bmi_p_values, data.frame(column = col, p_value = test_result))
}

# 4. Flag significant results
bmi_p_values <- bmi_p_values %>%
  mutate(significant = p_value < 0.05)

# 5. Print all p-values in formatted form
bmi_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# 6. Manual ANOVA check (optional)
summary(aov(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ bmi_group, data = merged_data_norm_bmi))

# 7. Append p-values as a row to the dataset
bmi_p_row <- merged_data_norm_bmi[nrow(merged_data_norm_bmi), ]
bmi_p_row[1, ] <- NA
bmi_p_row$bmi_group <- "p-value"

for (col in bmi_cols_to_test) {
  p_val <- bmi_p_values$p_value[bmi_p_values$column == col]
  if (length(p_val) > 0) {
    bmi_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

merged_data_with_bmi_p_values <- rbind(merged_data_norm_bmi, bmi_p_row)


# 8. Optional: Run post-hoc tests for significant variables
significant_vars <- bmi_p_values %>% filter(significant) %>% pull(column)

for (var in significant_vars) {
  cat("\nPairwise comparisons for:", var, "\n")
  formula <- as.formula(paste0("`", var, "` ~ bmi_group"))
  aov_model <- aov(formula, data = merged_data_norm_bmi)
  tukey <- TukeyHSD(aov_model)
  print(tukey)
}

# turn this tukey test into a table 

# 8.1 Create a summary table for Tukey's HSD results (significant only)
tukey_summary <- data.frame(
  variable = character(),
  comparison = character(),
  estimate = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (var in significant_vars) {
  formula <- as.formula(paste0("`", var, "` ~ bmi_group"))
  aov_model <- aov(formula, data = merged_data_norm_bmi)
  tukey_result <- TukeyHSD(aov_model)
  
  # Loop through comparisons and extract significant ones
  for (comparison in rownames(tukey_result[[1]])) {
    estimate <- tukey_result[[1]][comparison, "diff"]
    p_value <- tukey_result[[1]][comparison, "p adj"]
    
      tukey_summary <- rbind(tukey_summary, data.frame(
        variable = var,
        comparison = comparison,
        estimate = estimate,
        p_value = p_value
      ))
  }
}

# View the filtered significant results
print(tukey_summary)


# 9. Print last rows for verification
cat("\nLast rows of the dataset with BMI p-values row added:\n")
tail(merged_data_with_bmi_p_values, 5)

# 10. Output number of columns
cat("This is the number of columns:", ncol(merged_data_with_bmi_p_values), "\n")
```

BMI Charts 
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Get the significant columns with p-values and significance stars
sig_bmi_vars <- bmi_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    # Create more descriptive labels based on cell types
    short_name = case_when(
      grepl("Early NK", column) ~ "Early NK cells",
      grepl("CD4\\+ TPex", column) ~ "CD4+ TPex cells",
      TRUE ~ paste0("Cell Type ", LETTERS[seq_len(n())])
    ),
    # Keep original A, B labels but will add descriptions elsewhere
    short_label = paste0(LETTERS[seq_len(n())], " ", significance)
  )

# 2. Prepare long-format data with the significant variables
# Convert to percentages by multiplying by 100
long_data <- merged_data_norm_bmi %>%
  select(bmi_group, all_of(sig_bmi_vars$column)) %>%
  pivot_longer(cols = -bmi_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,  # Convert to percentage
    short_variable = factor(
      sig_bmi_vars$short_label[match(variable, sig_bmi_vars$column)],
      levels = sig_bmi_vars$short_label
    ),
    # Reorder BMI groups logically
    bmi_group = factor(bmi_group, levels = c("Average", "Overweight", "Obese"))
  )

# 3. Create the plot with Relative Abundance as the y-axis (now as percentage)
ggplot(long_data, aes(x = bmi_group, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, 
               fill = "black", color = "black") +
  facet_wrap(~ short_variable, scales = "free_y", ncol = 2) +
  # Add caption to explain what A and B represent
  labs(
    title = "Distribution of Cell Populations by BMI Group",
    subtitle = "Significant differences in cell frequency between BMI categories",
    x = "BMI Category",
    y = "Percentage of Live Cells (%)",
    caption = paste0(
      "A: Early NK cells (", sub(".*Live Cells/", "", sig_bmi_vars$column[1]), ")\n",
      "B: CD4+ TPex cells (", sub(".*Live Cells/", "", sig_bmi_vars$column[2]), ")"
    )
  ) +

  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 0.5),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(fill = "lightblue", color = NA),
    strip.text = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(hjust = 0, size = 10, face = "italic")
  )


# Optional output summary
cat("\nSignificant BMI-based variables:\n")
sig_output <- sig_bmi_vars %>%
  mutate(
    summary = paste0(short_name, " ", significance, ": ", 
                     column, 
                     " | p = ", sprintf("%.4g", p_value), 
                     " ", significance)
  ) %>%
  pull(summary)

cat(paste(sig_output, collapse = "\n"))

# how to do we find which weight category is significant for a given cell
# normalize the weights in the significant cell line 
# TUKEY TEST

```

# Summary from tukey test BMI

#Post hoc analysis using Tukey's HSD revealed a significant reduction in Early NK cells in the Overweight group compared to individuals with Average BMI (p = 0.030), while the Obese group did not differ significantly from either #Average (p = 0.322) or Overweight (p = 0.286) groups. Similarly, CD4+ TPex cell frequencies were significantly elevated in the Overweight group relative to the Average BMI group (p = 0.035), with no significant differences observed #between Overweight and Obese groups.

Athero observation 

```{r}
library(dplyr)
library(stringr)

# 1. Create athero_group column
merged_data_norm_athero <- merged_data_norm %>%
  mutate(athero_group = case_when(
    str_detect(`Procedure(s) - notes`, regex("CABG|CORONARY ARTERY BYPASS GRAFT", ignore_case = TRUE)) ~ "CABG",
    TRUE ~ "NON-CABG"
  )) %>%
  arrange(athero_group) %>%
  select(athero_group, everything())

# 2. Select numeric columns to test
athero_cols_to_test <- merged_data_norm_athero %>%
  select(where(is.numeric)) %>%
  colnames()

# 3. Initialize p-value table
athero_p_values <- data.frame(column = character(), p_value = numeric(), stringsAsFactors = FALSE)

# 4. Run t-tests for each numeric column
for (col in athero_cols_to_test) {
  group1 <- merged_data_norm_athero[merged_data_norm_athero$athero_group == "CABG", col]
  group2 <- merged_data_norm_athero[merged_data_norm_athero$athero_group == "NON-CABG", col]
  
  test_result <- tryCatch(t.test(group1, group2, var.equal = TRUE), error = function(e) NULL)
  
  if (!is.null(test_result)) {
    athero_p_values <- rbind(athero_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    athero_p_values <- rbind(athero_p_values, data.frame(column = col, p_value = NA))
  }
}

# 5. Add significance flag
athero_p_values <- athero_p_values %>%
  mutate(significant = p_value < 0.05)

# 6. Print formatted p-values
athero_p_values %>%
  arrange(p_value) %>%
  mutate(p_value = sprintf("%.8f", p_value)) %>%
  print(n = Inf)

# 7. Manual test (optional)
t.test(`FlowCut-passed/Cells/Single Cells/Live Cells/CD45+|count` ~ athero_group,
       data = merged_data_norm_athero,
       var.equal = TRUE)

# 8. Append p-values as a new row
athero_p_row <- merged_data_norm_athero[nrow(merged_data_norm_athero), ]
athero_p_row[1, ] <- NA
athero_p_row$athero_group <- "p-value"

for (col in athero_cols_to_test) {
  p_val <- athero_p_values$p_value[athero_p_values$column == col]
  if (length(p_val) > 0) {
    athero_p_row[[col]] <- sprintf("%.8f", p_val)
  }
}

merged_data_with_athero_p_values <- rbind(merged_data_norm_athero, athero_p_row)

# 9. Preview and save
cat("\nLast rows of the dataset with CABG p-values row added:\n")
tail(merged_data_with_athero_p_values, 5)

# 10. Export to CSV
```
Athero charts

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)

# 1. Get the significant variables from CABG t-tests
sig_athero_vars <- athero_p_values %>%
  filter(significant == TRUE) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*"
    ),
    short_name = LETTERS[seq_len(n())],
    short_label = paste0(short_name, " ", significance)
  )

# 2. Split into cell and non-cell variables
sig_cell_vars <- sig_athero_vars %>%
  filter(str_detect(column, "FlowCut")) %>%
  mutate(
    simplified_name = column %>%
      str_replace_all("FlowCut-passed/Cells/Single Cells/Live Cells/", "") %>%
      str_replace_all("\\|count", "") %>%
      str_replace_all("\\|freq_of_parent", "%")
  )

sig_noncell_vars <- sig_athero_vars %>%
  filter(!str_detect(column, "FlowCut")) %>%
  mutate(simplified_name = column)

# 3. Split cell variables into two groups for better visualization
n_cell_vars <- nrow(sig_cell_vars)
split_point <- ceiling(n_cell_vars / 2)

sig_cell_vars_1 <- sig_cell_vars[1:split_point, ]
sig_cell_vars_2 <- sig_cell_vars[(split_point + 1):n_cell_vars, ]

# 4. Create label maps for both groups
label_map_cell_1 <- setNames(sig_cell_vars_1$short_label, sig_cell_vars_1$column)
label_map_cell_2 <- setNames(sig_cell_vars_2$short_label, sig_cell_vars_2$column)
label_map_noncell <- setNames(sig_noncell_vars$short_label, sig_noncell_vars$column)

# 5. Long-format data for cell variables - Group 1
long_data_cell_1 <- merged_data_norm_athero %>%
  select(athero_group, all_of(sig_cell_vars_1$column)) %>%
  pivot_longer(cols = -athero_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,
    short_label = label_map_cell_1[variable]
  )

# 6. Long-format data for cell variables - Group 2
long_data_cell_2 <- merged_data_norm_athero %>%
  select(athero_group, all_of(sig_cell_vars_2$column)) %>%
  pivot_longer(cols = -athero_group, names_to = "variable", values_to = "value") %>%
  mutate(
    value = value * 100,
    short_label = label_map_cell_2[variable]
  )

# 7. Long-format data for non-cell variables
long_data_noncell <- merged_data_norm_athero %>%
  select(athero_group, all_of(sig_noncell_vars$column)) %>%
  pivot_longer(cols = -athero_group, names_to = "variable", values_to = "value") %>%
  mutate(
    short_label = label_map_noncell[variable]
  )

# 8. Create Plot 1 - First half of cell variables
athero_cell_plot_1 <- ggplot(long_data_cell_1, aes(x = athero_group, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "black", color = "black") +
  facet_wrap(~ short_label, scales = "free_y", ncol = 4) +
  labs(
    title = "Distribution of Cell Populations by Procedure Group (Part 1)",
    subtitle = "Significant differences in cell frequency between CABG and NON-CABG groups",
    x = "Procedure Group",
    y = "Percentage of Live Cells (%)",
    caption = paste(
      paste0(sig_cell_vars_1$short_label, ": ", sig_cell_vars_1$simplified_name),
      collapse = "\n"
    )
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(fill = "lightblue", color = NA),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic"),
    plot.margin = margin(t = 20, r = 20, b = 40, l = 20)
  )

# 9. Create Plot 2 - Second half of cell variables
athero_cell_plot_2 <- ggplot(long_data_cell_2, aes(x = athero_group, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "black", color = "black") +
  facet_wrap(~ short_label, scales = "free_y", ncol = 4) +
  labs(
    title = "Distribution of Cell Populations by Procedure Group (Part 2)",
    subtitle = "Significant differences in cell frequency between CABG and NON-CABG groups",
    x = "Procedure Group",
    y = "Percentage of Live Cells (%)",
    caption = paste(
      paste0(sig_cell_vars_2$short_label, ": ", sig_cell_vars_2$simplified_name),
      collapse = "\n"
    )
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(fill = "lightblue", color = NA),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic"),
    plot.margin = margin(t = 20, r = 20, b = 40, l = 20)
  )

# 10. Non-cell variable plot (unchanged, but with better spacing)
athero_noncell_plot <- ggplot(long_data_noncell, aes(x = athero_group, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5, color = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "black", color = "black") +
  facet_wrap(~ short_label, scales = "free_y", ncol = 4) +
  labs(
    title = "Distribution of Non-Cell Features by Procedure Group",
    subtitle = "Significant differences in clinical variables between CABG and NON-CABG groups",
    x = "Procedure Group",
    y = "Raw Values",
    caption = paste(
      paste0(sig_noncell_vars$short_label, ": ", sig_noncell_vars$simplified_name),
      collapse = "\n"
    )
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(fill = "lightblue", color = NA),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic"),
    plot.margin = margin(t = 20, r = 20, b = 40, l = 20)
  )

# 11. Display all plots
print(athero_cell_plot_1)
print(athero_cell_plot_2)
if(nrow(sig_noncell_vars) > 0) {
  print(athero_noncell_plot)
}

# 12. Optional: Run t-tests for all significant variables
cat("\n=== STATISTICAL TESTS ===\n")
for (i in 1:nrow(sig_athero_vars)) {
  sig_var <- sig_athero_vars$column[i]
  short_name <- sig_athero_vars$short_name[i]
  simplified <- sig_athero_vars$simplified_name[i]
  
  cat("\nStatistical test for", short_name, ":", simplified, ":\n")
  t_test_result <- t.test(as.formula(paste0("`", sig_var, "` ~ athero_group")), 
                          data = merged_data_norm_athero)
  print(t_test_result)
  cat("---\n")
}

# 13. Print comprehensive variable key
cat("\n=== VARIABLE KEY ===\n")
cat("CELL POPULATIONS (Part 1):\n")
for (i in 1:nrow(sig_cell_vars_1)) {
  cat(paste0(
    sig_cell_vars_1$short_label[i], ": ",
    sig_cell_vars_1$simplified_name[i],
    " (p = ", sprintf("%.4g", sig_cell_vars_1$p_value[i]), ")",
    "\n"
  ))
}

cat("\nCELL POPULATIONS (Part 2):\n")
for (i in 1:nrow(sig_cell_vars_2)) {
  cat(paste0(
    sig_cell_vars_2$short_label[i], ": ",
    sig_cell_vars_2$simplified_name[i],
    " (p = ", sprintf("%.4g", sig_cell_vars_2$p_value[i]), ")",
    "\n"
  ))
}

if(nrow(sig_noncell_vars) > 0) {
  cat("\nNON-CELL VARIABLES:\n")
  for (i in 1:nrow(sig_noncell_vars)) {
    cat(paste0(
      sig_noncell_vars$short_label[i], ": ",
      sig_noncell_vars$simplified_name[i],
      " (p = ", sprintf("%.4g", sig_noncell_vars$p_value[i]), ")",
      "\n"
    ))
  }
}
```




library(car)  # for leveneTest

smoking_p_values <- data.frame()

for (col in smoking_cols_to_test) {
  group1 <- merged_data_norm_smoking[merged_data_norm_smoking$smoking_group == "Smoker", col]
  group2 <- merged_data_norm_smoking[merged_data_norm_smoking$smoking_group == "Non-smoker", col]
  
  # Combine into one data frame for Levene's test
  test_data <- data.frame(
    value = c(group1, group2),
    group = rep(c("Smoker", "Non-smoker"), c(length(group1), length(group2)))
  )
  
  # Run Levene's Test
  levene_result <- tryCatch(leveneTest(value ~ group, data = test_data), error = function(e) NULL)
  equal_var <- TRUE
  
  if (!is.null(levene_result)) {
    p_lev <- levene_result$`Pr(>F)`[1]
    equal_var <- p_lev > 0.05  # If p > 0.05, assume equal variance
  }

  # Run t-test with appropriate variance assumption
  test_result <- tryCatch(t.test(group1, group2, var.equal = equal_var), error = function(e) NULL)

  if (!is.null(test_result)) {
    smoking_p_values <- rbind(smoking_p_values, data.frame(column = col, p_value = test_result$p.value))
  } else {
    smoking_p_values <- rbind(smoking_p_values, data.frame(column = col, p_value = NA))
  }
}



